---
title: "Data-analysis"
output: github_document
---

```{r}
library(tidyverse)
library(lubridate)
library(readr)
library(janitor)
library(patchwork)
library(scales)
library(moderndive)
library(stats)
library(sandwich)
```

For this project we are using two datasets available from the CDC. 

The first is a dataset that includes weekly estimates of expected deaths by week for each US states, as well as observed deaths. It is available for download here:
https://data.cdc.gov/NCHS/Excess-Deaths-Associated-with-COVID-19/xkkf-xrst/

```{r message=FALSE, warning=FALSE}
excess_deaths <- read_csv("data/Excess_Deaths_Associated_with_COVID-19.csv") %>% 
  clean_names() %>% 
  rename(end_week=week_ending_date) %>% 
  filter(type=="Predicted (weighted)"&outcome=="All causes") 
```    

The second dataset includes the weekly counts of COVID-19 related deaths in the for each US state. And is avalailable here: https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Week-Ending-D/r8kw-7aab

```{r message=FALSE, warning=FALSE}
covid_deaths <- read_csv("data/Provisional_COVID-19_Death_Counts_by_Week_Ending_Date_and_State.csv") %>% 
  clean_names() %>%
  mutate(end_week=mdy(end_week))%>%
  select(end_week,state,covid_19_deaths,total_deaths,percent_of_expected_deaths,pneumonia_deaths, pneumonia_and_covid_19_deaths, influenza_deaths, pneumonia_influenza_or_covid_19_deaths) 

merged_data <- inner_join(covid_deaths,excess_deaths) %>% 
  mutate(excess_non_covid=excess_higher_estimate-covid_19_deaths)


```

These datasets are updated frequently. This analysis is based on data downloaded on December 1, 2020. Also because death reporting takes time this analysis focuses on deaths reported through the end of October 2020.


To explore the relationship between excess deaths, deaths attributed to COVID-19, and excess deaths that were not attributed to COVID-19, we generated time series plots for these quanities for the entire nation. 

```{r message=FALSE, warning=FALSE}

panel1=merged_data %>% 
  filter(state=="United States") %>% 
   ggplot(aes(x=end_week,y=covid_19_deaths))+
   labs(x = "Date", y = "COVID-19 deaths") +
   geom_point(alpha=0.5, size=3)+
   geom_smooth(se=FALSE, span=0.2) +
    scale_x_date(
    labels = date_format("%m-%Y"),
    breaks= c(as.Date("2020/2/1"),as.Date("2020/5/1"),as.Date("2020/8/1"), as.Date("2020/11/1")),
    limits= c(as.Date("2020/2/1"), as.Date("2020/10/31")))+
  theme(axis.text.x=element_text(angle=45,hjust=1))+
   scale_y_continuous(limits=c(-1000, 25000))


panel2= merged_data  %>% 
  filter(state=="United States") %>% 
   ggplot(aes(x=end_week,y=excess_higher_estimate))+
   labs(x = "Date", y = "Excess deaths") +
   geom_point(alpha=0.5, size=3)+
   geom_smooth(se=FALSE, span=0.2) +

    scale_x_date(
    labels = date_format("%m-%Y"),
    breaks= c(as.Date("2020/2/1"),as.Date("2020/5/1"),as.Date("2020/8/1"), as.Date("2020/11/1")),
    limits= c(as.Date("2020/2/1"), as.Date("2020/10/31")))+
  theme(axis.text.x=element_text(angle=45,hjust=1))+
   scale_y_continuous(limits=c(-1000, 25000))

panel3=
 merged_data %>%
  filter(state=="United States") %>% 
  ggplot(aes(x=end_week,y=excess_non_covid))+
   labs(x = "Date", y = "Excess deaths not attributed to COVID-19") +
   geom_point(alpha=0.5, size=3)+
   geom_smooth(se=FALSE, span=0.2) +

    scale_x_date(
    labels = date_format("%m-%Y"),
    breaks= c(as.Date("2020/2/1"),as.Date("2020/5/1"),as.Date("2020/8/1"), as.Date("2020/11/1")),
    limits= c(as.Date("2020/2/1"), as.Date("2020/10/31")))+
  theme(axis.text.x=element_text(angle=45,hjust=1))+
   scale_y_continuous(limits=c(-1000, 25000))

panel2 + panel1 + panel3
   
```
This code chunk calculates the number of excess deaths, covid deaths, and excess non-covid deaths in the United States from February through October 2020.

```{r}
merged_data %>% 
  filter(state=="United States" & end_week < as.Date("2020-11-1")) %>% 
  summarise(
    excess_deaths=sum(excess_higher_estimate),
    covid_deaths=sum(covid_19_deaths),
    excess_not_covid=sum(excess_higher_estimate)-sum(covid_19_deaths))
```
Approximately 1/3 of excess deaths were not attributed to COVID.


Next I was interested in getting a sense of whether the rate of excess non-covid deaths varied across States.

In the next code chunk I aggregate the data by state over the entire period, merge the data with another data set containing state level population estimates for 2020. 

```{r}
states= merged_data %>% 
  group_by(state) %>% 
  filter(end_week < as.Date("2020-11-1") & state !="United States" & state!="New York City") %>% 
  filter(!is.na(covid_19_deaths)) %>% 
  filter(!is.na(excess_higher_estimate)) %>% 
  summarise(
    excess_deaths=sum(excess_higher_estimate),
    covid_deaths=sum(covid_19_deaths),
    excess_not_covid=sum(excess_higher_estimate)-sum(covid_19_deaths)) 

state_pop <- read_csv("data/State_Populations.csv") %>% 
  clean_names()

states=right_join(states,state_pop) %>% 
  mutate(excess_death_r=excess_deaths/population*100000) %>% 
  mutate(covid_death_r=covid_deaths/population*100000) %>% 
  mutate(excess_not_covid_r=excess_not_covid/population*100000)


```


```{r}

panel1 = states %>% 
 ggplot( aes(x=covid_death_r,y=excess_not_covid_r, size=population))+
   labs(x = "COVID-19 deaths \n per 100,000 residents", y = "Excess non-COVID-19 deaths \n per 100,000 residents") +
   geom_point(alpha=0.5)+
  geom_smooth(weight="population")+
  theme(legend.position = "none") +
  scale_y_continuous(limits=c(0, 300))


panel2 = states %>% 
 ggplot( aes(x=covid_death_r,y=excess_death_r, size=population))+
   labs(x = "COVID-19 deaths \n per 100,000 residents", y = "Excess non-COVID-19 deaths \n per 100,000 residents") +
   geom_point(alpha=0.5)+
  geom_smooth(, weight="population")+
 theme(legend.position = "none") +
  scale_y_continuous(limits=c(0, 300))

panel2 + panel1
```

Next we explore whether the duration of stay at home orders are associated with excess non-covid related mortality. We import and link data on the lengths of each states stay-at-home-orders based on information aggregated from by the National Academy of State Health Policy (https://www.nashp.org/governors-prioritize-health-for-all/). We join these data to our state level data set plot stay-at-home order duratoin vs non-covid-19 excess mortality rates. 

```{r}
stay_at_home_orders <- read_csv("data/stay_at_home_orders_final.csv", 
    col_types = cols(Start = col_date(format = "%m/%d/%Y"), 
        End = col_date(format = "%m/%d/%Y"))) %>% 
  clean_names() %>% 
  mutate(stay_at_home=if_else(length==0,FALSE,TRUE))

states=right_join(states,stay_at_home_orders)

states %>% 
 ggplot( aes(x=length,y=excess_not_covid_r, size=population))+
   labs(x = "Length of stay-at-home order (days)", y = "Excess non-COVID-19 deaths \n per 100,000 residents") +
   geom_point(alpha=0.5)+
  geom_smooth(, weight="population")+
 theme(legend.position = "none") +
  scale_y_continuous(limits=c(0, 100))+

```

```{r}

stay_at_home_model <- lm.cluster(excess_not_covid_r ~ length, data = states,cluster=state)
# Get regression table:
get_regression_table(stay_at_home_model)

```
