---
title: "COVID-19 Impacts"
output: 
  html_document: 
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE,fig.width=8, fig.height=6, out.width="100%")
library(tidyverse)
library(plotly)
library(readxl)
library(dplyr)
library(lubridate)
library(patchwork)
library(scales)
library(moderndive)
library(stats)
library(skimr)
library(kableExtra)
library(usmap)
library(viridis)
library(ggthemes)

options(scipen=10)  

```
<br>  

### Are there actually more deaths reported in 2020 than past years?  

To examine the issue, we use the data from 
```{r}
excess_deaths <- read_csv("data/Excess_Deaths_Associated_with_COVID-19.csv") %>% 
  janitor::clean_names() %>% 
  rename(end_week=week_ending_date) %>%
  filter(type=="Predicted (weighted)"&outcome=="All causes")

recorded_deaths = excess_deaths %>%
  separate(col=end_week,into=c("year","month","day"),sep="-") %>%
  select(c(year,month,state,observed_number)) %>%
  group_by(year,month,state)%>%
  summarise_each(funs(sum))%>%
  filter(state=="United States") 

recorded_deaths = recorded_deaths[-c(47),]


recorded_deaths_graph = recorded_deaths %>%
  ggplot(aes(x=month,y=observed_number,color=year)) + 
  geom_point() + geom_line(aes(group=year)) + 
  labs(
    title = "United States Total Recorded Deaths By Months",
    x = "Month",
    y = "Recorded Deaths"
  ) + 
  scale_color_viridis(discrete = TRUE, option = "C") +
  theme_tufte()

ggplotly(recorded_deaths_graph)
```

```{r}
recorded_deaths_bar = recorded_deaths %>%
  filter(month != 11 & month != 12) %>%
  select(year,month,observed_number) %>%
  group_by(year)%>%
  summarise(
    total_death = sum(observed_number)
  ) %>%
  ggplot(aes(x=year,y=total_death,fill=year)) + 
  geom_bar(stat="identity") +
  labs(
    title = "United States Total Reported Deaths of First 10 Months",
    x = "Year",
    y = "Recorded Deaths"
  ) +
  theme_tufte() +
  scale_fill_viridis(discrete = TRUE, option = "C") 


ggplotly(recorded_deaths_bar)
```
Yes, the following graph shows unequivocally that deaths reported since March 2020 has exceeded the past three years by a substantial margin. In the month of May, the recorded deaths number of 2020 is almost 50% higher than the average from 2017 to 2019.  


### Maybe reported deaths number was going to peak in 2020 due to demographic/environmental or other external variables anyways. Is the reported death spike not expected?
```{r}
excess_deaths_2 = excess_deaths %>% 
  separate(col=end_week,into=c("year","month","day"),sep="-") %>%
  mutate(death_diff = observed_number - upper_bound_threshold) %>%
  filter(state=="United States") %>%
  select(c(year,month,day,state,observed_number,upper_bound_threshold,death_diff)) %>%
  mutate(date = str_c(year,month,day,sep="-")) %>%
  mutate(date = as.Date(date)) %>%
  mutate(death_diff_percent = death_diff/upper_bound_threshold) %>%
  relocate(date,.before=state) %>%
  select(-c(month,day))

excess_deaths_2 = excess_deaths_2[-c(201),]

excess_deaths_graph = excess_deaths_2 %>%
  ggplot(aes(x=date,y=death_diff,color=year)) + 
  geom_point() + 
  labs(
    title = "Weekly Excess Deaths From All Causes",
    x = "Month",
    y = "Recorded Deaths - Upper Bound Projected Deaths "
  ) +
  scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
  scale_color_viridis(discrete = TRUE, option = "C") +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) 


ggplotly(excess_deaths_graph)
```

No, the spike in reported deaeth in 2020 is not expected at all according to the CDC weekly deaths projections. Over the past four years, aside from the winter of 2017 when recorded deaths number briefly eclipsed the upperbound deaths projections, recorded deaths number has been consistently at a lower level than the upperbound deaths projections. In 2020, the excess deaths number shot way past the projection as well as historical averages.  

### Okay, but how do I know this spike in reported deaths is directly correlated with COVID-19 mortalities?

```{r}
covid_deaths <- read.csv("data/Provisional_COVID-19_Death_Counts_by_Week_Ending_Date_and_State.csv") %>% 
  janitor::clean_names() %>%
  mutate(end_week=mdy(end_week))%>%
  select(end_week,state,covid_19_deaths,total_deaths,percent_of_expected_deaths,pneumonia_deaths, pneumonia_and_covid_19_deaths, influenza_deaths, pneumonia_influenza_or_covid_19_deaths) 

excess_deaths = excess_deaths %>%
  mutate(end_week=as.Date(end_week))

merged_data = inner_join(covid_deaths,excess_deaths) %>%
  filter(state=="United States" & end_week < as.Date("2020-11-1"))

covid_v_excess = merged_data %>% 
 ggplot(aes(x=covid_19_deaths,y=excess_higher_estimate))+
   labs(x = "Weekly Reported COVID-19 deaths ", 
        y = "Weekly Reported Excess deaths",
        title = "Weekly Reported COVID-19 Deaths vs. Weekly Excess Deaths"
       ) +
   geom_point(alpha=0.5)+
  geom_smooth(se=FALSE)+ 
  theme_tufte()


ggplotly(covid_v_excess)
```

```{r}
reg_1 = lm(excess_higher_estimate ~ covid_19_deaths, data=merged_data)
reg_1 %>%
  broom::tidy() %>%
  select(term,estimate,p.value)%>%
  kable(digits = 10)
```

By mapping reported COVID-19 deaths against weekly excess deaths, there is a clear, significant linear relationship. In other words, the change of weekly excess deaths can almost be perfectly explained by change in COVID-19 deaths. Therefore, we can conclude that COVID-19 is indeed the culprit of the increased excess deaths of 2020.

### How much of the excess deaths increase can be attributed to Covid-19?

```{r}
merged_data = merged_data %>%
  mutate(non_covid_excess = excess_higher_estimate-covid_19_deaths)

pie_data = merged_data %>%
  select(non_covid_excess,covid_19_deaths) %>%
  summarize_if(is.numeric, sum, na.rm=TRUE) %>%
  pivot_longer(
    non_covid_excess:covid_19_deaths,
    names_to = "type",
    values_to = "deaths"
  )

pie_graph = pie_data %>%
  ggplot(aes(x="", y=deaths,fill=type)) +
  geom_bar(stat="identity", width=1,color="white") +
  coord_polar("y", start=0) + 
  theme_void() +
  scale_fill_manual(values=c("navy", "maroon")) +
  labs(
    title = "Composition of Total Excess Deaths"
  )

pie_graph
```

```{r}
deaths_comp_df = merged_data %>%
  select(end_week,non_covid_excess,covid_19_deaths,excess_higher_estimate) %>%
  rename(excess_deaths = excess_higher_estimate) %>%
  rename(non_covid_excess_deaths = non_covid_excess) %>%
   reshape::melt(id.vars = c("end_week"), measure.vars = c("non_covid_excess_deaths","covid_19_deaths","excess_deaths"))

deaths_comp_graph = deaths_comp_df %>%
      rename(death_type = variable) %>%
  ggplot(aes(x=end_week,y=value,color=death_type)) + 
  geom_point() + geom_line(aes(group=death_type)) + 
  labs(
    title = "United States Monthly Recorded Deaths By Types",
    x = "Month",
    y = "Recorded Deaths"
  )+
  theme_tufte()

ggplotly(deaths_comp_graph)
```

```{r}
covid_v_non = merged_data %>% 
 ggplot(aes(x=covid_19_deaths,y=non_covid_excess))+
   labs(x = "Weekly Reported COVID-19 deaths ", 
        y = "Weekly Reported non-Covid Excess deaths",
        title = "Weekly Reported COVID-19 Deaths vs. non-Covid Excess Deaths"
       ) +
   geom_point(alpha=0.5)+
  geom_smooth(method = "lm")+ 
  theme_tufte()

ggplotly(covid_v_non)
```

### What states have relatively the largest increase in excess death this year?
```{r}
extra_deaths_state = excess_deaths %>%
  filter(end_week > as.Date("2020-01-01")) %>%
  select(end_week,state,observed_number,average_expected_count,year) %>%
  mutate(deaths_diff = observed_number - average_expected_count,
         deaths_diff_percent = (deaths_diff/average_expected_count * 100))

extra_deaths_state = extra_deaths_state %>%
  select(state,deaths_diff_percent) %>%
  drop_na(deaths_diff_percent) %>%
  group_by(state)%>%
  summarise_each(funs(mean)) %>%
  mutate(deaths_diff_percent = round(deaths_diff_percent,digits=2)) %>%
  filter(state != "New York City" & state != "District of Columbia" & state != "United States")

map_graph_1 = plot_usmap(data = extra_deaths_state, values = "deaths_diff_percent") + 
    scale_fill_viridis(direction = -1) + 
  theme(legend.position = "right") +
  labs(fill = "Excess deaths per 100 expected deaths",
       title = "Excess Deaths vs. Expected Deaths By States")

ggplotly(map_graph_1)
```
  
### What States will have the highest excess deaths per 100,000 people by the end of 2020?
```{r}
state_pop = read_csv("data/State_Populations.csv") %>% 
  janitor::clean_names()

excess_deaths_3 = excess_deaths %>%
  filter(state != "United States" & state != "New York City" & end_week < as.Date("2020-3-1")) %>%
  select(state,total_excess_higher_estimate_in_2020) %>%
  group_by(state)%>%
  summarise(
    excess_estimate = mean(total_excess_higher_estimate_in_2020)
  )


us_map = inner_join(excess_deaths_3,state_pop) %>%
  mutate(excess_death_per = excess_estimate/population*100000)

map_graph = plot_usmap(data = us_map, values = "excess_death_per", color = "black") + 
    scale_fill_viridis(direction = -1) + 
  theme(legend.position = "right") +
  labs(fill = "Excess deaths per 100,000 people",
       title = "Excess Deaths Rate By States")

 ggplotly(map_graph)

```