---
title: "Prevention Outcomes"
output: 
  html_document: 
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE,fig.width=8, fig.height=6, out.width="100%")

library(tidyverse)
library(plotly)
library(readxl)
library(dplyr)
library(lubridate)
library(patchwork)
library(scales)
library(moderndive)
library(stats)
library(skimr)
library(kableExtra)
library(usmap)
library(janitor)
library(miceadds)
library(stringr)
library(viridis)
library(ggthemes)


options(scipen=10)

```

### How have states responded to the pandemic?

To evaluate the relationship between prevention measures and excess deaths, we first pulled data from the [National Academy of State Health Policy](https://www.nashp.org/governors-prioritize-health-for-all/) with specifications of each states' steps in mitigating the spread of COVID-19. We then mapped each state's lockdown duration, mask mandate, and travel restrictions.  

```{r}
excess_deaths <- read_csv("data/Excess_Deaths_Associated_with_COVID-19.csv") %>% 
  clean_names() %>% 
  rename(end_week=week_ending_date) %>% 
  filter(type=="Predicted (weighted)"&outcome=="All causes") 

covid_deaths <- read_csv("data/Provisional_COVID-19_Death_Counts_by_Week_Ending_Date_and_State.csv") %>% 
  clean_names() %>%
  mutate(end_week=mdy(end_week))%>%
  select(end_week,state,covid_19_deaths,total_deaths,percent_of_expected_deaths,pneumonia_deaths, pneumonia_and_covid_19_deaths, influenza_deaths, pneumonia_influenza_or_covid_19_deaths) 

merged_data <- inner_join(covid_deaths,excess_deaths) %>% 
  mutate(excess_non_covid=excess_higher_estimate-covid_19_deaths)

states= merged_data %>% 
  group_by(state) %>% 
  filter(end_week < as.Date("2020-11-1") & state !="United States" & state!="New York City") %>% 
  filter(!is.na(covid_19_deaths)) %>% 
  filter(!is.na(excess_higher_estimate)) %>% 
  summarise(
    excess_deaths=sum(excess_higher_estimate),
    covid_deaths=sum(covid_19_deaths),
    excess_not_covid=sum(excess_higher_estimate)-sum(covid_19_deaths)) 

state_pop <- read_csv("data/State_Populations.csv") %>% 
  clean_names()

states=right_join(states,state_pop) %>% 
  mutate(excess_death_r=excess_deaths/population*100000) %>% 
  mutate(covid_death_r=covid_deaths/population*100000) %>% 
  mutate(excess_not_covid_r=excess_not_covid/population*100000)

stay_at_home_orders <- read_csv("data/stay_at_home_orders_final.csv", 
    col_types = cols(Start = col_date(format = "%m/%d/%Y"), 
        End = col_date(format = "%m/%d/%Y"))) %>% 
  clean_names() %>% 
  mutate(stay_at_home=if_else(length==0,FALSE,TRUE))

states=right_join(states,stay_at_home_orders)

preventions_df = read_csv("data/prevention_measures.csv") %>% 
  clean_names() %>% 
  mutate(
    mask_mandate = str_sub(mask_mandate, 0, 3),
    mask_mandate = ifelse(mask_mandate == "No", FALSE, TRUE),
    travel_restrictions = ifelse(travel_restrictions == "There are no statewide restrictions.", FALSE, TRUE)
    ) 

states_df=right_join(states,preventions_df)

prevent_map1 = plot_usmap(data = states_df, values = "length", color = "black") +
  scale_fill_viridis(direction = -1) + 
  theme(legend.position = "right")+
  labs(fill = "Duration of Lockdown",
       title = "Length of States' Stay-at-Home Orders")


ggplotly(prevent_map1)

```


```{r}
map2_title = list(
  text = "State has mask mandate",
  xref = "paper",
  yref = "paper",
  yanchor = "bottom",
  xanchor = "center",
  align = "center",
  x = 0.5,
  y = 0.95,
  showarrow = FALSE)

map3_title = list(
  text = "State has travel restrictions",
  xref = "paper",
  yref = "paper",
  yanchor = "bottom",
  xanchor = "center",
  align = "center",
  x = 0.5,
  y = 0.95,
  showarrow = FALSE)

prevent_map2 = plot_usmap(data = states_df, values = "mask_mandate", color = "black") + 
  theme(legend.position = "none")+
  scale_colour_viridis_d(option = "inferno")

prevent_map3 = plot_usmap(data = states_df, values = "travel_restrictions", color = "black")+
  theme(legend.position = "right")+
  scale_colour_viridis_d(option = "inferno")

prevent_map2 = ggplotly(prevent_map2) %>% layout(annotations = map2_title)
prevent_map3 = ggplotly(prevent_map3) %>% layout(annotations = map3_title)

subplot(style(prevent_map2, showlegend = F), prevent_map3)
```


**Answer:** The response to the COVID-19 pandemic has differed from state to state. Generally, the Great Plains states took the fewest prevention measures with never issuing a shelter-in-place order, mask mandate, or travel restrictions. Overall, with no clear federal direction and guidance on prevention measures, the United States experienced a fragemented response in the first six months of the pandemic. 

### Do "lockdowns kill"? What is the association between lockdown durations and excess deaths not related to COVID-19.

To investigate whether states that required residents to stay at home experienced a higher burden of other deaths outside of COVID-19, we combined data from [National Academy of State Health Policy](https://www.nashp.org/governors-prioritize-health-for-all/) with *Excess Deaths Associated with COVID-19* from the [Centers for Disease Control and Prevention](https://data.cdc.gov/NCHS/Excess-Deaths-Associated-with-COVID-19/xkkf-xrst/). After filtering out excess deaths attributed to COVID-19, we fitted a linear regression model with the length of stay-at-home order as a continuous exposure and number of excess deaths not attributed to COVID-19 per 100,000 as the outcome.   

```{r, collapse=TRUE}
states %>% 
 ggplot( aes(x=length,y=excess_not_covid_r, size=population))+
   labs(x = "Length of stay-at-home order (days)", y = "Excess COVID-19 deaths not attributed\n to COVID-19 per 100,000 residents") +
  geom_point(alpha=0.5)+
  geom_smooth(, weight="population")+
 theme(legend.position = "none") +
  scale_y_continuous(limits=c(0, 100))

mod3 = lm.cluster(excess_not_covid_r ~ length, data = states, cluster="state")

mod3_sum = 
  summary(mod3)%>% 
  as_tibble()

mod3_CI = 
  confint(mod3)%>% 
  as_tibble()


mod3_df = 
  bind_cols(mod3_sum, mod3_CI)%>% 
  mutate(
    Variables = c("Intercept", "Length of stay-at-home order")
  ) %>% 
  relocate(Variables) %>% 
  kbl(caption="Linear model output assessing association of length of lockdown and non-COVID-19 excess deaths", digits = 2) %>%
  kable_classic()

mod3_df
```

**Answer:** No. We can see from this model that there is no statistically significant association between length of lockdowns and excess deaths not related to COVID-19. These results undermine the assertion that "lockdowns kill people". While there may be above average deaths in the U.S. that are not due to COVID-19, it is evident that lockdowns are not a major driver of them. 

<br><br>

### Well, how about masks and travel restrictions? Did states that decreased civil liberties see a significant increase in excess deaths not related to COVID-19?  

To address a far-fetched assertion that imposing on the civil liberties of state citizens may lead to an above average number of deaths not related to COVID-19, we used the same combined data with excess deaths and prevention measures. We then fitted two linear models with binary exposure indicating the presence of travel restrictions or a mask mandate and made number of excess deaths not attributed to COVID-19 per 100,000 as the outcome. 

```{r, collapse=TRUE}

mod4 = lm.cluster(excess_not_covid_r ~ travel_restrictions, data = states_df, cluster="state")


mod4_sum = 
  summary(mod4) %>% 
  as_tibble()

mod4_CI = 
  confint(mod4) %>% 
  as_tibble()

mod4_df = 
  bind_cols(mod4_sum, mod4_CI) %>% 
  mutate(
    Variables = c("Intercept", "Travel restrictions")
  ) %>% 
  relocate(Variables) %>% 
  kbl(caption="Linear model output assessing association of travel restrictions and excess non-COVID-19 deaths per 100,000", digits = 2) %>%
  kable_classic()

mod4_df
```


```{r, collapse = TRUE}
mod5 = lm.cluster(excess_not_covid_r ~ mask_mandate, data = states_df, cluster="state")


mod5_sum = 
  summary(mod5) %>% 
  as_tibble()

mod5_CI = 
  confint(mod5) %>% 
  as_tibble()

mod5_df = 
  bind_cols(mod5_sum, mod5_CI) %>% 
  mutate(
    Variables = c("Intercept", "Mask mandate")
  ) %>% 
  relocate(Variables) %>% 
  kbl(caption="Linear model output assessing association of mask mandates and excess non-COVID-19 deaths per 100,000", digits = 2) %>%
  kable_classic()

mod5_df
```

**Answer:** Neither mask mandates nor travel restrictions had significant relationships with the number of above average deaths not related to COVID-19. 

<br><br>

### Is there an association between a states' lockdown duration and non-COVID-19 deaths at home?

Next, we wanted to see if states with longer lockdown durations had more non-COVID-19 deaths occur at home. To do this, we pulled data from the *Provisional COVID-19 Death Counts by Place of Death and State* from the [Centers for Disease Control and Prevention](https://data.cdc.gov/NCHS/Provisional-COVID-19-Death-Counts-by-Place-of-Deat/uggs-hy5q) and combined it with the *Excess Deaths Associated with COVID-19* and prevention measures datasets from earlier. After filtering for deaths not attributed to COVID-19 which occurred in the household, we plotted non-COVID deaths at home per 100,000 persons against the length of stay-at-home orders. We then fitted a linear model assessing the association of length of lockdown and non-COVID-19 deaths at home per 100,000 residents.

```{r, collapse=TRUE}
home_deaths = read_csv("data/place_of_death.csv") %>% 
  clean_names() %>% 
  filter(place_of_death == "Decedent's Home") %>% 
  filter(state != "United States") %>%
  filter(state != "New York City") %>% 
  filter(state != "Puerto Rico") %>% 
  mutate(non_covid_home_deaths = total_deaths-covid19_deaths) %>% 
  select(state, non_covid_home_deaths)

states_final = 
  right_join(states, preventions_df) %>%
  right_join(home_deaths) %>% 
  mutate(non_covid_home_deaths_r = non_covid_home_deaths/population*100000)



states_final %>% 
 ggplot( aes(x=length,y=non_covid_home_deaths_r, size=population))+
   labs(x = "Length of stay-at-home order (days)", y = "Deaths at home not attributed\n to COVID-19 per 100,000 residents") +
  geom_point(alpha=0.5)+
  geom_smooth(, weight="population")+
 theme(legend.position = "none") +
  scale_y_continuous(limits=c(150, 450))

mod6 = lm.cluster(non_covid_home_deaths_r ~ length, data = states_final, cluster="state")

mod6_sum = 
  summary(mod6) %>% 
  as_tibble()

mod6_CI = 
  confint(mod6) %>% 
  as_tibble()

mod6_df = 
  bind_cols(mod6_sum, mod6_CI) %>% 
  mutate(
    Variables = c("Intercept", "Length of lockdown")
  ) %>% 
  relocate(Variables) %>% 
  kbl(caption="Linear model output assessing association of length of lockdown and non-COVID-19 deaths at home per 100,000 residents", digits = 4) %>%
  kable_classic()

mod6_df
```

**Answer:** No. We do not see any significant relationship between how long state residents were asked to shelter-in-place and the number of non-COVID-19 deaths at home per 100,000 persons. This evidence weakens claims that requesting citizens to stay at home to prevent the spread of COVID-19 results in excess deaths at home from other causes   

### After implementation, were prevention measures effective in decreasing total excess deaths?

We lastly wanted to evaluate if prevention measures that states took later infleunced the amount of above average deaths that state experienced. To do this analysis, we once again used our merged data of *Excess Deaths Associated with COVID-19*, state population, and prevention measures. We focused our investigation on excess deaths that occurred after the first peak and after lockdown measures were lifted to see if there were lasting protections from prevention measures. We fit a linear model clustering by state with duration of lockdown, mask mandates, and travel restrictions against total excess deaths per capita. 


```{r, collapse=TRUE}
post_prevent = merged_data %>% 
  group_by(state) %>% 
  filter(end_week > as.Date("2020-6-1") & state !="United States" & state!="New York City") %>% 
  filter(!is.na(covid_19_deaths)) %>% 
  filter(!is.na(excess_higher_estimate)) %>% 
  summarise(
    excess_deaths=sum(excess_higher_estimate),
    covid_deaths=sum(covid_19_deaths),
    excess_not_covid=sum(excess_higher_estimate)-sum(covid_19_deaths)) 

post_prevent_final = 
  right_join(post_prevent, preventions_df) %>% 
  right_join(stay_at_home_orders) %>% 
  right_join(state_pop) %>% 
  mutate(excess_death_r=excess_deaths/population*100000) %>% 
  mutate(covid_death_r=covid_deaths/population*100000) %>% 
  mutate(excess_not_covid_r=excess_not_covid/population*100000)

mod7 = lm.cluster(excess_death_r ~ length  + mask_mandate + travel_restrictions, data = post_prevent_final, cluster="state")

mod7_sum = 
  summary(mod7) %>% 
  as_tibble()

mod7_CI = 
  confint(mod7) %>% 
  as_tibble()

mod7_df=
  bind_cols(mod7_sum, mod7_CI) %>% 
  mutate(
    Variables = c("Intercept", "Length of lockdown", "Mask mandate", "Travel restrictions")
  ) %>% 
  relocate(Variables) %>% 
  kbl(caption="Linear model output assessing association of prevention measures and excess deaths per 100,000 residents post first peak (after June 1st, 2020)", digits = 4) %>%
  kable_classic()

mod7_df
```

**Answer:** Each additional day of lockdown did decrease the number of excess deaths per capita. Additionally, states that imposed travel restrictions did decrease excess deaths per capita by 25 after the first peak compared to those that did not. Overall, it is difficult to measure the association between excess deaths and these prevention measures because of unaccounted confounding relationships, reverse causation, and other unmeasured variables. 

